A Systematic Literature Review on Extensions to the Role-Based Access Control Reference Model

### 2014-02-23 paper/sections/abstract.tex
Formatting updates.

\noindent
{\bf Context}: Since the United States National Institute of Standards and Technology proposed Role-Based Access Control (RBAC) as documented in the RBAC Reference Model in the late 1990's, domain-targeted extensions have been proposed.  For example, the mobile domain has identified a need for permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding how RBAC extension models are evaluated by providing an assessment of the state of RBAC extension models.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review of RBAC extension models that began with 1,716 papers of which 27 were deemed as primary sources for inclusion.
\\
\\
\noindent
{\bf Results}: We identified and classified the RBAC extension models into eight extension categories: Constraint, Context, Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. Only 8 of the 27 papers provided an implementation of their model in the form of an enterprise application or prototype. The primary domains that inspired extensions were the medical domain with 9 of the 27 models, enterprise workflows with 5 of 27, and mobile computing with 5 of 27. The state of evaluation of RBAC model extensions is inconsistent with 4 papers comparing themselves to the RBAC Reference Model, 14 presenting example scenarios, and 12 lacking any evaluation that we could discern. 
\\
\\
\noindent
{\bf Conclusions}: Our literature review shows that all eight of the RBAC extension categories we identified deal with context whereby the privileges provided to a role are environmental depending upon factors such as location or time. 
The magnitude and scope of extensions to the RBAC standard suggests a revised standard may be beneficial.

>>>>

\noindent
{\bf Context}: Since the United States National Institute of Standards and Technology proposed Role-Based Access Control (RBAC) as documented in the RBAC Reference Model in the late 1990's, domain-targeted extensions have been proposed.  For example, the mobile domain has identified a need for permission granting and role activation.
\\
\\
\noindent
{\bf Objective}: The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding how RBAC extension models are evaluated by providing an assessment of the state of RBAC extension models.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review of RBAC extension models that began with 1,716 papers of which 27 were deemed as primary sources for inclusion.
\\
\\
\noindent
{\bf Results}: We identified and classified the RBAC extension models into eight extension categories: Constraint, Context, Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. Only 8 of the 27 papers provided an implementation of their model in the form of an enterprise application or prototype. The primary domains that inspired extensions were the medical domain with 9 of the 27 models, enterprise workflows with 5 of 27, and mobile computing with 5 of 27.
\\
\\
\noindent
{\bf Conclusions}: Our literature review shows that all eight of the RBAC extension categories we identified deal with context whereby the privileges provided to a role are environmental depending upon factors such as location or time. 
RBAC extension model evaluation lacks a consistent set of metrics and evaluation of current models was found to range from providing scenario examples of the model in action to comparison to the RBAC standard.
The magnitude and scope of extensions to the RBAC standard suggests a revised standard may be beneficial but in the meantime this work can serve as a starting place for researchers and practitioners.

### 2014-02-23 paper/sections/analysis.tex
Address comments given by Dr. Williams
Updates.
Formatting updates.

The primary source of validation a developer or practitioner may encounter is a qualitative discussion of real-world
scenarios and how the proposed model can tackle those situations. In rare cases, five primary sources, the model 
authors provide some discussion of how the RBAC Reference Model is deficient in tackling the scenario.
In Table \ref{tab:evaluations}, we observe that one paper includes a case study to examine how the proposed model works in practice. Discussions
of how an extension model handles a real-world scenario provides developers and practitioners anecdotal evidence at best
for what types of situations the proposed model could handle. Further, by not providing a comparison to the RBAC Reference Model, developers may be left implementing a more complex model to address their requirements when the RBAC Reference Model would
have sufficed.  Further, given the nature of access control models as grounded in application to enterprise implementations, 
case studies of a model in action provide developers with evidence that the model works as intended when applied.

>>>>

The primary source of validation a developer or practitioner may encounter is a qualitative discussion of real-world
scenarios and how the proposed model can tackle those situations. 
For five of the primary sources \cite{bao08:role} \cite{zou2009crbac} \cite{zhang06:collaborative} \cite{zhao2008flexible} \cite{ray07:spatio} the model authors provide some discussion of how the RBAC Reference Model is deficient in tackling the scenario.
In Table \ref{tab:evaluations}, we observe that one paper includes a case study to examine how the proposed model works in practice. Discussions
of how an extension model handles a real-world scenario provides developers and practitioners anecdotal evidence at best
for what types of situations the proposed model could handle. Further, by not providing a comparison to the RBAC Reference Model, developers may be left implementing a more complex model to address their requirements when the RBAC Reference Model would
have sufficed.  Further, given the nature of access control models as grounded in application to enterprise implementations, 
case studies of a model in action provide developers with evidence that the model works as intended when applied.

### 2014-02-23 paper/sections/discussion.tex
Formatting updates.

The research questions we identified, presented results for and analyzed provide specific views of portions of the RBAC extension model landscape. 
By stepping back, and looking across all the research questions we can arrive at cross-cutting concerns and identify areas that may benefit from future work.
We break the discussion into two issues: recommendations for the enhancement of the evaluation of RBAC extension models to role-based access control models and guidelines for developers needing to choose an RBAC extension model to meet their requirements.

@@

A small percentage of primary sources presented evaluations or implementations of their model. This result leaves researchers and practitioners little data from which to compare one model to the next. Given the wide swath of popular and critical domains being targeted by RBAC extension models, Hu and Scarfone proposed metrics for access control system evaluation or a security system in general~\cite{hu2012:NIST7874}. They collect metrics, which can be applicable for not only access control models, but also requirements, system implementation, and extended application such as verification and testing. For example, ``Ease of Privilege Assignments'' metric measures the number of steps for assigning, changing, or deleting a privilege for users and roles when a user or administrator manages RBAC. This metric can be applicable for not only the RBAC Reference Model but also RBAC extension models.
%Future work could include providing metric measurements for the primary sources presented in this paper. 
We plan to use these metrics proposed by Hu and Scarfone for any extension models presented in this paper. 
A developer could attempt to apply these metrics to an RBAC extension model not already evaluated when attempting to choose a model. 
These metrics focus on comparing specific features support of access control models.

@@

To adopt a model in practice, software developers would implement the model in real system environments based on intended use of the models. Typically, a model provides abstract formal representation of RBAC extension and its operation. To configure intended use (i.e., which subjects can access which resources), we define an access control policy, which specifies high-level rules such as which subjects can access to which resources. Given an access control model and policy, security mechanism provides low-level implementation in a system, which controls access of roles.

@@

This standard does not provide a specific guideline for implementation of various RBAC extension models.
Moreover, when software developers implemented RBAC extension models, they require clear security requirements of extended role-based access control, which identifies security objectives, intended environments, and assumed threats. Moreover, software developers understand how this RBAC extension model overlaps with other access control models when more than two access control models are integrated into a single system. A survey~\cite{o20102010} shows that over 50\% of users at organizations with more than 500 employees are given some of their permissions to access resources based on RBAC. To adopt RBAC into a system, organizations often use hybrid approaches, which combine RBAC and access control lists because specific user types, systems, and workflow may not be effective to manage access based on roles. Therefore, when RBAC is extended and combined with other access control models, software developers often require to provide a formal representation, which can support of proof of the model. Then, software developers implement RBAC extension models based on NIST standard of RBAC implementation. Especially, software developers understand intended use of RBAC extension models to meet a new system requirement. For example, for a spatial RBAC model, software developers incorporate additional spatial constraints, which can be either static or dynamic, in practice into the RBAC model.

>>>>

The research questions we identified, presented results for and analyzed provide a view of the RBAC extension model landscape. 
By looking across all the research questions we can arrive at cross-cutting concerns and identify areas that may benefit from future work.
We break the discussion into two issues: recommendations for the enhancement of the evaluation of RBAC extension models to role-based access control models and guidelines for developers needing to choose an RBAC extension model to meet their requirements.

@@

A small percentage of primary sources presented robust evaluations or implementations of their model. 
This result leaves researchers and practitioners little data from which to compare one model to the next. 
Given the wide swath of popular and critical domains being targeted by RBAC extension models, Hu and Scarfone proposed metrics for access control system evaluation or a security system in general~\cite{hu2012:NIST7874}. 
They collect metrics, which can be applicable for not only access control models, but also requirements, system implementation, and extended application such as verification and testing. 
For example, ``Ease of Privilege Assignments'' metric measures the number of steps for assigning, changing, or deleting a privilege for users and roles when a user or administrator manages RBAC. 
This metric can be applicable for not only the RBAC Reference Model but also RBAC extension models.
We propose the metrics introduced by Hu and Scarfone should be applied to the extension models presented in this paper as well as any future models.
A developer could attempt to apply these metrics to an RBAC extension model not already evaluated when attempting to choose a model. 
These metrics focus on comparing feature support of access control models.

@@

To adopt a model in practice, software developers would implement the model in real system environments based on intended use of the models. 
As shown by RQ6, all primary sources were found to provide an abstract formal representations of their extension model and the model's operation.
The presence of an abstract formal representation of the model stands as a starting place for developers to configure intended use (i.e. which subjects can access which resources), and define an access control policy (i.e. which specifies high-level rules such as which subjects can access to which resources).

@@

This standard does not provide a specific guideline for implementation of various RBAC extension models.
Moreover, when software developers implemented RBAC extension models, they require clear security requirements of extended role-based access control, which identifies security objectives, intended environments, and assumed threats. 
Moreover, software developers understand how this RBAC extension model overlaps with other access control models when more than two access control models are integrated into a single system. 
A survey~\cite{o20102010} shows that over 50\% of users at organizations with more than 500 employees are given some of their permissions to access resources based on RBAC. 
To adopt RBAC into a system, organizations often use hybrid approaches, which combine RBAC and access control lists because specific user types, systems, and workflow may not be effective to manage access based on roles. 
Therefore, when RBAC is extended and combined with other access control models, software developers can provide a formal representation which can serve as support for proof of the model. 
Then, software developers implement RBAC extension models based on NIST standard of RBAC implementation. 
Especially, software developers understand intended use of RBAC extension models to meet a new system requirement. 
For example, for a spatial RBAC model, software developers incorporate additional spatial constraints, which can be either static or dynamic, in practice into the RBAC model.

### 2014-02-23 paper/sections/intro.tex
Formatting updates.

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item RQ1: How can RBAC extension models be classified?
\item RQ2: What are motivations behind the RBAC extension models?
\item RQ3: How are RBAC extension models implemented and evaluated by their authors?
\item RQ4: What domains have been targeted by RBAC extension models?
\item RQ5: What commonalities exist across RBAC extension models?
\end{itemize}

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item A starting place for researchers looking to tackle new problems in the realm of authorization and as a reference guide for discovering
the current state of the RBAC extension models. 
\item A basis for comparison and look up for what extension model to use for developers looking to find a model to fit their access control needs.
\item A summary of current evaluation methods used in research on RBAC extension models.
\end{itemize}

>>>>

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item RQ1: How can RBAC extension models be classified?
\item RQ2: What are motivations behind the RBAC extension models?
\item RQ3: How are RBAC extension models implemented by their authors?
\item RQ4: How are RBAC extension models evaluated by their authors?
\item RQ5: What domains have been targeted by RBAC extension models?
\item RQ6: What commonalities exist across RBAC extension models?
\end{itemize}

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item A starting place for researchers in the realm of authorization and as a reference guide for discovering the current state of the RBAC extension models. 
\item A basis for comparison and look up for what extension model to use for developers looking to find a model to fit their access control needs.
\item A summary of current evaluation methods used in research on RBAC extension models.
\end{itemize}
